---
title: "Gapminder"
author: "Trevor McDonald"
date: 2026-09-28
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder%>%glimpse()
```

**Observations**:

- We have country, continent, year, life expectancy, population, and gdp per captia.

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()
year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
library(ggrepel)
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min,
    gdpPercap > 100000) %>%
  ggplot(aes(pop, gdpPercap, color = country)) +
  geom_point() +
  geom_label_repel(
    aes(label = country, fill = country)
  )
gapminder %>%
  filter(
    year == year_min,
    gdpPercap < 100000
    ) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot()
  
```

**Observations**:

- Oceania displays the highest average GDP per capita, but it also has very few data points to draw from.
- Europe has the second highest average GDP per capita but it also has the most variation.
- Each continent besides oceania has a few outliers that are higher than the average, but none that are lower than the average.

**Difficulties & Approaches**:

- One thing I noticed was that there was a very high outlier value that made it very hard to find observations otherwise. To solve, this, I isolated that data point to see which country it was. It turned out to be Kuwait, and when I looked up Kuwaits gdp percapita was in 1952, I found that it wasn't even tracked until 1962. This means I could filter it out from my graph.
- I chose a boxplot because boxplots give me a lot more information to look at rather than just raw points. Seeing each quartile allowed me to make more observations about trends and which values are truly outliers.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
library(ggrepel)
## TASK: Identify the outliers from q2
outlier_africa <- gapminder %>% filter(continent == "Africa", gdpPercap > 3000, year == year_min)
outlier_americas <- gapminder %>% filter(continent == "Americas", gdpPercap > 7500, year == year_min)
outlier_asia <- gapminder %>% filter(continent == "Asia", gdpPercap > 5000 & gdpPercap < 100000, year == year_min)
outlier_europe <- gapminder %>% filter(continent == "Europe", gdpPercap > 12500, year == year_min)
combined <- rbind(outlier_africa, outlier_americas, outlier_asia, outlier_europe)
combined
```

**Observations**:

- Identify the outlier countries from q2
  - I found th outliers for each continent shown in the box plots.
  Africa: Angola, Gabon, South Africa
  Americas: Canada, United States, Venezuela
  Asia: Bahrain, Saudi Arabia
  Europe: Switzerland

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(
    year == year_max | year == year_min,
    gdpPercap < 100000) %>%
  ggplot(aes(continent, gdpPercap, group = str_c(year, continent))) +
  geom_boxplot(
    aes(fill = year),
    position = "dodge"
  ) + 
  geom_point(
    data = . %>% filter(
      country %in% c("Angola", "Gabon", "South Africa", "Canada", "United States", "Venezuela", "Bahrain", "Saudi Arabia", "Switzerland"),
      ),
    mapping = aes(color = country),
    size = 2,
    position = position_dodge(width = 0.75)
  )

```

**Observations**:

- All continents had an average increase in gdp per capita for the 25th, 50th, and 75th quartiles
- Oceania's box is a lot larger. I looked it up on google and this is because countries like New Zealand or pacific islands didn't start tracking their gdp per capita until after 1952.
- In Africa there are now more outliers, but the bottom 50% today is the closest to the bottom 50% in 1952 for Africa than all other continents.
- Venezuela is no longer an outlier in the Americas. Canada and U.S. sit well above any other country in the Americas
- Asia has a relatively small 25 - 50 quartile compared to its 50 -75 quartile. 
- Asia seems to exhibit the largest range followed by Europe.

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>% 
  filter(
    year == year_max) %>%
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(aes(size = pop, color = continent), alpha = 0.7) +
  geom_smooth(se = FALSE)
```

- This graph isn't trying to answer any question in-particular, it's to try and see general trends and see if any of them are interesting.
- My initial thought was to see if population had anything to do with the gdp per capita, but looking at this graph, there's nothing particularly interesting to me about population that sticks out.
- Visually, Africa exhibits a much lower gdp per capita and life expectancy than other continents
- Life expectancy seems to have a logarithmic relationship to gdp per capita.
- There seems to be a sharp increase in life expectancy with increasing gdp per capita, but that seems to diminish around 10,000 gdp per capita.

```{r q5-task2}
## TASK: Your second graph
quartiles <- gapminder %>%
  group_by(year) %>%
  summarize(
    bottom = quantile(lifeExp, 0),
    quart1 = quantile(lifeExp, .25),
    quart2 = quantile(lifeExp, .5),
    quart3 = quantile(lifeExp, .75),
    top = quantile(lifeExp, 1)
  ) %>%
  pivot_longer(
    names_to = "quartile",
    values_to = "lifeExp",
    cols = -year
  )
quartiles
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  geom_boxplot(aes(group = year)) +
  geom_point(aes(color = continent)) +
  geom_line(data = quartiles, aes(linetype = quartile), color = "blue", size = 1)

gapminder %>%
  filter(lifeExp < 25)
  
```

- This graph tries to see if there is a difference in the life expectancy growth between different quartiles.
- The outlier in 1992 is Rwanda. At face value this can be explained by their civil war ranging from 1990-1993, but according to a quick google search the number of people it killed before 1992 was in the thousands, maybe tens of thousands. However, Rwanda's population at that time was north of 7 million. This suggests that this could be inaccurate data, and a google search supports that fact.
- There doesn't seem to be a major disparity in growth, except both the 0% line and the 25% line plateau after the 90s. There doesn't seem to be an easy answer to support why that is.
- The median has experienced the most growth out of all the quartiles. Life expectancy has gone up from ~45 to ~72.
- African countries are overly represented in the bottom 2 quartiles.

```{r q5-task3}
## TASK: Your third graph
lower <- gapminder %>%
  filter(continent == "Africa") %>%
  group_by(year) %>%
  summarize(
    country  = country[which.min(lifeExp)],
    lifeExp  = min(lifeExp),
  )
higher <- gapminder %>%
  filter(continent == "Africa") %>%
  group_by(year) %>%
  summarize(
    country = country[which.max(lifeExp)],
    lifeExp = max(lifeExp)
  )
higher
lower
library(ggrepel)
gapminder %>%
  filter(continent == "Africa",
         lifeExp > 25) %>%
  ggplot(aes(lifeExp, year)) +
  geom_boxplot(aes(group = year)) +
  geom_point(data = lower, color = "red") +
  geom_point(data = higher, color = "blue") +
  geom_text_repel(data = higher, aes(label = country), size = 3) +
  geom_text_repel(data = lower, aes(label = country), size = 3)
```

- Since Africa being grouped together in the bottom 50% of our data for our other two graphs, I wanted to take a little bit of a closer look.
- The two countries that appear at the top of life expectancy for Africa are Reunion and Mauritius which are both tiny islands off the coast of Madagascar. Both of these populations are around 1 million, and they seem to be more of a vacation spot than anything else.
- The countries on the lower end are gambia, Sierra Leone, Angola, Rwanda, Zambia, and Swaziland (Eswatini). I wanted to see if there was some geographical trend between the lowest life expectancy but I didn't see any.
- Another cause I hypothesized was the decolonization period that led to a lot of civil wars could explain the difference, but all of these countries had civil wars around the same time (1960s-1970s), so that wouldn't explain why the plateau has continued to today.


