---
title: "Vis: Perceptual Basics"
author: Zach del Rosario
date: 2020-08-06
output: github_document
time: 20
reading: 40
---

# Vis: Perceptual Basics

*Purpose*: Creating a *presentation-quality* graph is an exercise in *communication*. In order to create graphs that other people can understand, we should know some stuff about *how humans see data*. Through the required "reading" (video) you'll learn about visual perception, then put these ideas to use criticizing some graphs. Later, you'll use these ideas to *improve* some graphs.

*Reading*: [How Humans See Data](https://www.youtube.com/watch?v=fSgEeI2Xpdc&list=PLluqivwOH1ouKkbM0c6x-g7DQnXF0UmC0&index=37&t=0s) (Video)

```{r setup}
library(tidyverse)
```

## Criticize these graphs!
<!-- -------------------------------------------------- -->

Using the ideas from the reading (video), state some issues with the following graphs. As a reminder, the *visual hierarchy* is:

1. Position along a common scale
2. Position on identical but nonaligned scales
3. Length
4. Angle; Slope (With slope not too close to 0, $\pi/2$, or $\pi$.)
5. Area
6. Volume; Density; Color saturation
7. Color hue

### __q1__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q1-vis}
## NOTE: No need to edit; run and criticize
mpg %>%
  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

**Observations**:

- What are some issues with this graph?
- This graph uses length, as its main showcase feature. This is an issue because it's hard to tell the difference in the spread of the various car companies.
- It's hard to see if there's a general trend in this graph because the data is grouped in such a way that it's not showing any information.
- They're organized alphabetically which is almost never useful, including this case.

### __q2__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q2-vis}
## NOTE: No need to edit; run and criticize
as_tibble(mtcars) %>%
  mutate(model = rownames(mtcars)) %>%

  ggplot(aes(x = "", y = "", size = mpg)) +
  geom_point() +
  facet_wrap(~model)
```

- What are some issues with this graph?
- It's hard to tell the difference in between some points. For example, I have no idea whether the Datsun 710 has a better mpg or the Merc 230. Size is low on the visual heiarchy.

### __q3__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q3-vis}
## NOTE: No need to edit; run and criticize
diamonds %>%
  ggplot(aes(clarity, fill = cut)) +
  geom_bar()
```

- What are some issues with this graph?
- stacking data is never a good idea. It relies on length to tell the difference between some data. For example, it's really hard to tell the difference in the number of premium cut diamonds for the groups SI1 and VS2. Position along a common axis would be much better.

### __q4__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q4-vis}
## NOTE: No need to edit; run and criticize
diamonds %>%
  ggplot(aes(x = "", fill = cut)) +
  geom_bar() +
  coord_polar("y") +
  labs(x = "")
```

- What are some issues with this graph?
- It's a pie chart... But it relies on angle which is low on our visual heiarchy.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis07-perception-assignment.Rmd).
